A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing

### 2012-05-20 docs/paper.tex
Fix spelling errors

\section{Introduction}
In this project we aim to analyze and improve the performance of Python’s dictionary or dict.  Python is an extremely popular general purpose computing language for several reasons:
\begin{enumerate}
 \item Clean, elegant near-pseudo code syntax leads to clear code 
 \item Widely distributed and available
 \item Large quantity of external libraries for specific features 
\end{enumerate}

@@

\subsection{Motivation}
Python heavily utilizes dictionary data structure for several of its internal language features. This extensive internal use of dictionaries means that a running Python program has many dictionaries active simultaneously, even in cases where the the user’s program code does not explicitly use a dictionary.

@@

In this section we give a brief overview of the key implementation details of
dictionaries in CPython 2.7.3.  We should note that Python’s dictionary supports
several different data types as keys. However as we noted earlier, the
dictionaries underlying class instances and modules have only strings as keys.
Therefore optimizing a dictionary which have string-only keys is crucial to the
runtime of any Python program. CPython accommodates for this optimization by
changing its dictionary lookup method as necessary. First when a dictionary is
instantiated, CPython uses a string-only lookup function until a search for
non-string is requested, in this case CPython falls back to a more general
lookup method and uses this general function onward. There are two main
optimizations in the string-only lookup function.  First, since the string
comparisons do not raise any exceptions by design, some of the error checking
logic can been removed. Second, the string-only lookup function can skip the
rich set of comparisons ( $<=$, $>=$, $<$, $>$, $==$, $!=$) which arbitrary data
types can provide since string type does not have these special cases.  As it
can be seen, string-only keys can dominate the runtime of a dictionary and
CPython uses some measures to optimize for this common case. Because of this
importance we mostly limit ourselves to the study of new implementations of
hashing the string type for the rest of the paper.

@@

After implementation and profiling, we found that dictionary table lookups had
over twice as many level 1 cache misses with tabulation hashing than without.
Hypothesizing that the tabulation tables were evicting other, more relavant
data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
instructions into our hash code to tell the CPU's caching system that it should
deprioritize storing it in cache.  We used GCC's
\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
to do this.  However, this only yielded about a 1\% performance improvement.

@@

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Figure ~\ref{fig:tables} shows the performance of different numbers of
tabulation tables.  Eight turned out to be the sweetspot
 \begin{figure}[H]
   \centering
   \includegraphics[width=4in]{tables.pdf}
   \caption{6000000 keys}
   \label{fig:tables}
 \end{figure}

Eight tables provided the best performance.
\subsection{Benchmarks}
In order to analyze the result of this new dictionary we measured three main metrics. First, we compared the performance of the simple tabulation hashing with Python’s current hashing scheme on random strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
Because of the heavy utilization of dictionaries in Python internals one of the valuable benchmarks is to see how the new dictionary performs on them. We used PyBench benchmark suit for this task. We saw significant improvements of around \%20 in performance of some of the internals such as the special and normal class attribute lookups and string predicates. However, at the same time some of the other internals such as CPython exception handler were running around \%12 slower compared to the current implementation. On average a \%4.5 improvement in runtime was measures for running the whole benchmark suite.

\begin{table}
  \caption{Performance results on selected components of the PyBench suite.  We
    have ommitted the least significant performace results in terms of percent
    change.
  }
\begin{tabular}{| l | c |}
  \hline
  \textbf{Test Name} & \textbf{Performance Difference} \\ \hline
  Test Name & Performance Difference \\ \hline
  NormalClassAttribute &  19.20\% \\ \hline
  SpecialClassAttribute &  18.88\% \\ \hline
  NormalInstanceAttribute &  17.77\% \\ \hline
  StringPredicates &  15.25\% \\ \hline
  UnicodePredicates &  13.82\% \\ \hline
  SimpleListManipulation &  13.03\% \\ \hline
  CompareUnicode &  11.27\% \\ \hline
  SimpleComplexArithmetic &  10.83\% \\ \hline
  SpecialInstanceAttribute &  10.52\% \\ \hline
  TryFinally &  10.35\% \\ \hline
  CreateUnicodeWithConcat &  8.52\% \\ \hline
  StringSlicing &  6.52\% \\ \hline
  Small Performance Changes Ommitted & ... \\ \hline
  UnicodeMappings &  -1.27\% \\ \hline
  ListSlicing &  -2.40\% \\ \hline
  ComplexPythonFunctionCalls &  -3.55\% \\ \hline
  TupleSlicing &  -3.72\% \\ \hline
  WithRaiseExcept &  -5.89\% \\ \hline
  TryRaiseExcept &  -11.99\% \\ \hline
  Totals & 4.54\% \\ \hline
\end{tabular}
\end{table}

@@

Different request rates (up to 2000 requests per second) were tested on this sample server and their corresponding response rates were calculated using httpef, a HTTP performance measurement tool. Considering the variance of request rates on different iterations of the test, we could not find any significant improvement of the response rate of the new implementation. We found that on the average performance was similar however the new implementation showed a smaller standard deviation in the response rates. This finding followed by other benchmarks showed that the new implementation gives more consistent runtime on different iterations of a same code compared to the current CPython implementation. This consistency is a desired behavior. We suspect that this behavior is a result of the randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=12cm]{appsbench.pdf}
  \caption{Prcentage of overall improvements in some real-world Python applications. Nearly none of the applications performed worse. Python Pickle and html5lib faster around \%3 to \%4. }
\end{figure}

>>>>

\section{Introduction}
In this project we aim to analyze and improve the performance of Python's dictionary or dict.  Python is an extremely popular general purpose computing language for several reasons:
\begin{enumerate}
 \item Clean, elegant near-pseudo code syntax leads to clear code 
 \item Widely distributed and available
 \item Large quantity of external libraries for specific features 
\end{enumerate}

@@

\subsection{Motivation}
Python heavily utilizes dictionary data structure for several of its internal language features. This extensive internal use of dictionaries means that a running Python program has many dictionaries active simultaneously, even in cases where the user's program code does not explicitly use a dictionary.

@@

In this section we give a brief overview of the key implementation details of
dictionaries in CPython 2.7.3.  We should note that Python's dictionary supports
several different data types as keys. However as we noted earlier, the
dictionaries underlying class instances and modules have only strings as keys.
Therefore optimizing a dictionary which have string-only keys is crucial to the
runtime of any Python program. CPython accommodates for this optimization by
changing its dictionary lookup method as necessary. First when a dictionary is
instantiated, CPython uses a string-only lookup function until a search for
non-string is requested, in this case CPython falls back to a more general
lookup method and uses this general function onward. There are two main
optimizations in the string-only lookup function.  First, since the string
comparisons do not raise any exceptions by design, some of the error checking
logic can been removed. Second, the string-only lookup function can skip the
rich set of comparisons ( $<=$, $>=$, $<$, $>$, $==$, $!=$) which arbitrary data
types can provide since string type does not have these special cases.  As it
can be seen, string-only keys can dominate the runtime of a dictionary and
CPython uses some measures to optimize for this common case. Because of this
importance we mostly limit ourselves to the study of new implementations of
hashing the string type for the rest of the paper.

@@

After implementation and profiling, we found that dictionary table lookups had
over twice as many level 1 cache misses with tabulation hashing than without.
Hypothesizing that the tabulation tables were evicting other, more relevant
data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
instructions into our hash code to tell the CPU's caching system that it should
deprioritize storing it in cache.  We used GCC's
\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
to do this.  However, this only yielded about a 1\% performance improvement.

@@

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Figure ~\ref{fig:tables} shows the performance of different numbers of
tabulation tables.  Eight turned out to be optimal.
 \begin{figure}[H]
   \centering
   \includegraphics[width=4in]{tables.pdf}
   \caption{6000000 keys}
   \label{fig:tables}
 \end{figure}

Eight tables provided the best performance.

 \begin{figure}[H]
   \centering
   \includegraphics[width=4in]{randstring.pdf}
   \caption{Random string performance}
   \label{fig:randstring}
 \end{figure}
 \begin{figure}[H]
   \centering
   \includegraphics[width=4in]{seqstring.pdf}
   \caption{Sequential string performance}
   \label{fig:seqstring}
 \end{figure}

\subsection{Benchmarks}
In order to analyze the result of this new dictionary we measured three main metrics. First, we compared the performance of the simple tabulation hashing with Python’s current hashing scheme on random strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
Because of the heavy utilization of dictionaries in Python internals one of the valuable benchmarks is to see how the new dictionary performs on them. We used PyBench benchmark suit for this task. We saw significant improvements of around \%20 in performance of some of the internals such as the special and normal class attribute lookups and string predicates. However, at the same time some of the other internals such as CPython exception handler were running around \%12 slower compared to the current implementation. On average a \%4.5 improvement in runtime was measures for running the whole benchmark suite.

\begin{table}
  \caption{Performance results on selected components of the PyBench suite.  We
    have omitted the least significant performance results in terms of percent
    change.
  }
\begin{tabular}{| l | c |}
  \hline
  \textbf{Test Name} & \textbf{Performance Difference} \\ \hline
  NormalClassAttribute &  19.20\% \\ \hline
  SpecialClassAttribute &  18.88\% \\ \hline
  NormalInstanceAttribute &  17.77\% \\ \hline
  StringPredicates &  15.25\% \\ \hline
  UnicodePredicates &  13.82\% \\ \hline
  SimpleListManipulation &  13.03\% \\ \hline
  CompareUnicode &  11.27\% \\ \hline
  SimpleComplexArithmetic &  10.83\% \\ \hline
  SpecialInstanceAttribute &  10.52\% \\ \hline
  TryFinally &  10.35\% \\ \hline
  CreateUnicodeWithConcat &  8.52\% \\ \hline
  StringSlicing &  6.52\% \\ \hline
  Small Performance Changes Omitted & ... \\ \hline
  UnicodeMappings &  -1.27\% \\ \hline
  ListSlicing &  -2.40\% \\ \hline
  ComplexPythonFunctionCalls &  -3.55\% \\ \hline
  TupleSlicing &  -3.72\% \\ \hline
  WithRaiseExcept &  -5.89\% \\ \hline
  TryRaiseExcept &  -11.99\% \\ \hline
  Totals & 4.54\% \\ \hline
\end{tabular}
\end{table}

@@

Different request rates (up to 2000 requests per second) were tested on this
sample server and their corresponding response rates were calculated using
httperf, a HTTP performance measurement tool. Considering the variance of request rates on different iterations of the test, we could not find any significant improvement of the response rate of the new implementation. We found that on the average performance was similar however the new implementation showed a smaller standard deviation in the response rates. This finding followed by other benchmarks showed that the new implementation gives more consistent runtime on different iterations of a same code compared to the current CPython implementation. This consistency is a desired behavior. We suspect that this behavior is a result of the randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=12cm]{appsbench.pdf}
  \caption{Percentage of overall improvements in some real-world Python applications. Nearly none of the applications performed worse. Python Pickle and html5lib faster around \%3 to \%4. }
\end{figure}
